---
title: "Usage"
output:
  rmarkdown::html_document:
    toc: true
    toc_title: "Content"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  eval      = TRUE,
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 8,
  fig.align = "center",
  dpi = 150,
  out.width = "100%"
)
```

```{r, include = FALSE}
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
library(pak)
pak::pkg_install(
  c(
    "dplyr",
    "sf",
    "mapview",
    "viridis",
    "spatialData"
  ),
  ask = FALSE
)
```

```{r}
library(dplyr)
library(sf)
library(mapview)
library(viridis)
library(spatialData)
library(spatialFolds)
```

```{r}
data("vi")
data("interaction")
data("trees")
```


# Cast functions

These functions cast POINT `sf` dataframes and `xy` data frames.

## `cast_df_to_sf()`

Transforms a data frame with coordinate columns to a POINT `sf` dataframe.

```{r}
#build xy dataframe
vi_xy <- vi |>
  sf::st_coordinates() |>
  as.data.frame()

head(vi_xy)
```

```{r}
vi_sf <- spatialFolds::cast_df_to_sf(
  df = vi_xy
)

vi_sf
```

## `cast_sf_to_xy()`

Extracts a two-column `xy` matrix from an `sf` dataframe. For POLYGON geometries, centroids are used.

```{r}
vi_xy_matrix <- cast_sf_to_xy(vi)
head(vi_xy_matrix)
```

```{r}
# Also works on polygons (uses centroids)
trees_xy <- cast_sf_to_xy(trees)
head(trees_xy)
```

## `cast_sf_to_bbox()`

Extracts the bounding box of an `sf` object as a one-row rectangular polygon `sf` dataframe.

```{r}
interaction_bbox <- cast_sf_to_bbox(interaction)

plot(sf::st_geometry(interaction))
plot(sf::st_geometry(interaction_bbox), add = TRUE, border = "red", lwd = 2)
```

## `cast_xy_to_xyz()`

Converts longitude/latitude coordinates to 3D Cartesian coordinates on a unit sphere. This eliminates discontinuities at the dateline and poles.

```{r}
vi_xyz <- cast_xy_to_xyz(vi_xy_matrix[1:5, ])
round(vi_xyz, 3)
```

```{r}
# North pole convergence: all longitudes map to the same point
pole_xy <- cbind(x = c(0, 90, 180), y = c(90, 90, 90))
round(cast_xy_to_xyz(pole_xy), 3)  # all rows identical at (0, 0, 1)
```


# Utility functions

## `utils_needs_spherical()`

Analyzes coordinate ranges (in degrees) to determine whether spherical geometry should be used. Returns `TRUE` for global data near the dateline or poles.

```{r}
# vi is in EPSG:4326 degrees and spans globally — returns TRUE
utils_needs_spherical(vi_xy_matrix)
```

```{r}
# Local geographic coordinates in degrees — returns FALSE
local_xy <- cbind(x = runif(100, -5, 5), y = runif(100, 40, 45))
utils_needs_spherical(local_xy)
```

```{r}
# Compute interaction_xy for use in later examples
# Note: interaction is in EPSG:25830 (UTM meters)
interaction_xy <- cast_sf_to_xy(interaction)
```


# Spatial Thinning

## `spatial_thinning()`

Resamples points to enforce minimum distance separation or reach a target sample size.

```{r}
# Auto-calculated distance
vi_thinned_auto <- spatial_thinning(df = vi)
nrow(vi_thinned_auto)
```

```{r}
# Fixed distance (degrees)
vi_thinned_dist <- spatial_thinning(df = vi, distance = 5)
nrow(vi_thinned_dist)
```

```{r}
# Target number of points
vi_thinned_target <- spatial_thinning(df = vi, target = 300)
nrow(vi_thinned_target)
```

## `thinning_to_distance()`

C++ function applying greedy sequential thinning with a fixed minimum distance. Returns indices of retained points.

```{r}
# interaction is in UTM meters; use 500 m as minimum separation distance
retained <- thinning_to_distance(xy = interaction_xy, distance = 500)
length(retained)
```

## `thinning_to_target()`

C++ function applying iterative thinning using binary search on distance until the target count is reached. Returns indices of retained points.

```{r}
retained_t <- thinning_to_target(xy = interaction_xy, target = 200L)
length(retained_t)
```


# Spatial Folds

## `spatial_folds()`

Generates training/testing folds for spatial cross-validation. Returns a data frame with one logical column per fold.

```{r}
# All three methods, single training fraction
folds <- spatial_folds(
  df = interaction,
  methods = c("contiguous", "random", "blocks"),
  training_fraction = 0.75,
  repetitions = 5,
  seed = 1
)
dim(folds)
names(folds)
```

```{r}
# Multiple training fractions
folds_multi <- spatial_folds(
  df = interaction,
  methods = "random",
  training_fraction = c(0.75, 0.5),
  repetitions = 3,
  seed = 1
)
names(folds_multi)
```

## `deduplicate_folds()`

Removes highly correlated (redundant) folds within each method-fraction group.

```{r}
folds_many <- spatial_folds(
  df = interaction,
  methods = "contiguous",
  training_fraction = 0.75,
  repetitions = 20,
  seed = 1
)
ncol(folds_many)
```

```{r}
folds_deduped <- deduplicate_folds(folds_many)
ncol(folds_deduped)
```

## `spatial_fold_plot()`

Visualizes one fold from a `spatial_folds()` result, showing training and testing samples.

```{r}
spatial_fold_plot(
  df = interaction,
  training_fold = folds$contiguous_0.75_1
)
```


# Low-level fold methods

## `training_mask()`

Thin R wrapper that routes to the appropriate C++ method and returns a single logical training/testing vector.

```{r}
mask_random <- training_mask(
  xy = interaction_xy,
  method = "random",
  seed = 42L,
  target = 750L
)
sum(mask_random)
```

## `method_contiguous_planar()`

C++ function for contiguous training folds using planar (rectangular) geometry. Suitable for local to subcontinental data.

```{r}
x_range <- diff(range(interaction_xy[, "x"]))
y_range <- diff(range(interaction_xy[, "y"]))

fold_planar <- method_contiguous_planar(
  xy = interaction_xy,
  center = 1L,
  step_x = x_range / 1000,
  step_y = y_range / 1000,
  target = 750L
)
sum(fold_planar)
```

## `method_contiguous_spherical()`

C++ function for contiguous training folds using spherical (angular cap) geometry. Required for global data near the dateline or poles.

```{r}
vi_xy_matrix <- cast_sf_to_xy(vi)
vi_xyz <- cast_xy_to_xyz(vi_xy_matrix)
x_range_vi <- diff(range(vi_xy_matrix[, "x"]))
y_range_vi <- diff(range(vi_xy_matrix[, "y"]))
ang_step <- mean(c(x_range_vi, y_range_vi)) / 1000 * pi / 180

fold_sph <- method_contiguous_spherical(
  xyz = vi_xyz,
  center = 1L,
  angular_step = ang_step,
  target = 5000L
)
sum(fold_sph)
```

## `method_random()`

C++ function for random data splitting using Fisher-Yates shuffle. Reproducible with a seed.

```{r}
fold_rnd <- method_random(
  xy = interaction_xy,
  seed = 42L,
  target = 750L
)
sum(fold_rnd)
```

## `method_blocks()`

C++ function that selects entire grid blocks until the target count is reached. Requires pre-computed block IDs from the internal `block_ids()` function.

```{r}
blocks_config <- validate_arg_blocks(blocks = NULL, df = interaction)
block_id_vec <- spatialFolds:::block_ids(interaction, blocks_config$rows, blocks_config$cols)

fold_blk <- method_blocks(
  xy = interaction_xy,
  block_id = block_id_vec,
  seed = 42L,
  target = 750L
)
sum(fold_blk)
```


# Validation functions

## `validate_arg_sf()`

Validates `sf` or `data.frame` inputs, converting data frames to sf when needed.

```{r}
validated <- validate_arg_sf(df = interaction)
names(validated)   # returns the validated sf object
```

## `validate_arg_xy()`

Validates xy matrix inputs. Returns the input silently on success; errors on failure.

```{r}
validate_arg_xy(xy = interaction_xy)  # silent on success
```

## `validate_arg_blocks()`

Validates the `blocks` argument and computes grid dimensions for the blocks method.

```{r}
blocks_auto <- validate_arg_blocks(blocks = NULL, df = interaction)
blocks_auto  # $rows, $cols
```

```{r}
blocks_manual <- validate_arg_blocks(blocks = c(5L, 10L), df = interaction)
blocks_manual
```

## `validate_arg_distance()`

Validates the `distance` argument for `spatial_thinning()`. Auto-calculates distance when `NULL`.

```{r}
dist_auto <- validate_arg_distance(df = interaction, distance = NULL)
dist_auto
```

## `validate_arg_target()`

Validates the `target` argument for `spatial_thinning()`.

```{r}
target_val <- validate_arg_target(df = interaction, target = 500L)
target_val
```
